Reasoning Graphically about Quantum Computation

### 2008-02-24 papers/reasoning-graphically/paper.tex
more stuff abut quantum computayion
git-svn-id: https://isaplanner.svn.sourceforge.net/svnroot/isaplanner/trunk/quantomatic@1358 f8f375fb-4f10-0410-8d47-b6fbdc7ca053

%%%%  use this to include graphics in the right place.
\newcommand{\inlinegraphic}[2]{
  %% todo -- make this thing calculate the height 
  %% itself based on a global scaling factor
  \dimendef\grafheight=255\dimendef\grafvshift=254
  \grafheight=#1
  \grafvshift=-0.5\grafheight
  \advance\grafvshift by 0.5ex
  \raisebox{\grafvshift}{\includegraphics[height=\grafheight]{#2}\xspace}
}

@@

Few examples of interesting compact closed categories are freely
generated.  In practice we work with categories presented as
generators and \emph{equations}.  For 
representing quantum computations, Coecke and Duncan
\cite{Coecke:2008jo} propose the following collection of generators:
\begin{description}
\item[Objects] the self dual object $Q$;
\item[Arrows] Two families of arrows:
  \begin{align}
  &\epsilon_Z : Q \to I, &\qquad&  \epsilon_X : Q \to I,\\
  &\delta_Z : Q \to Q \otimes Q, && \delta_X : Q \to Q \otimes Q, \\
  &\alpha_Z : Q \to Q &&  \alpha_X : Q \to Q   
  \end{align}
  where $\alpha \in [0,2\pi)$, and in addition $H:Q\to Q$.
\end{description}
We represent these two families, and their adjoints, by colours in our
graphical notation:
\begin{gather*}
  \epsilon_Z = \inlinegraphic{1.5em}{epsilon} \qquad
  \delta_Z = \inlinegraphic{1.5em}{delta} \qquad
  \epsilon_Z^\dag = \inlinegraphic{1.5em}{epsilondag} \qquad
  \delta_Z^\dag = \inlinegraphic{1.5em}{deltadag} \qquad
  \alpha_Z = \inlinegraphic{1.5em}{alpha} \qquad
\\  
  \epsilon_X = \inlinegraphic{1.5em}{redepsilon} \qquad
  \delta_X = \inlinegraphic{1.5em}{reddelta} \qquad
%  \epsilon_X^\dag = \inlinegraphic{1.5em}{redepsilondag} \qquad
%  \delta_X^\dag = \inlinegraphic{1.5em}{reddeltadag} \qquad
  \alpha_X = \inlinegraphic{1.5em}{redalpha} \qquad
\end{gather*}
The $H$ arrow is denoted by \inline
Each family $(\delta, \epsilon, \alpha)$ is governed by a set of
equations defining a forms a \emph{classical structure}
\cite{Coecke2006Quantum-Measure,Coecke2006POVMs-and-Naima} with a
\emph{phase group}.  The requisite equations state that $(\delta,
\epsilon )$ form a cocommutative comonoid:
\begin{gather}
  (\delta \otimes \id{Q}) \circ \delta 
   =   ( \id{Q} \otimes \delta) \circ \delta  
  \tag{A}\label{comonoid-assoc}
  \\
  (\epsilon \otimes \id{Q}) \circ \delta 
  = (\id{Q} \otimes \epsilon)\circ \delta 
  = \id{Q} \tag{U} \label{comonoid-unit}
  \\
  \delta = \sigma \circ \delta. \tag{C}\label{comonoid-comm}
\end{gather}
Their adjoints $(\delta^\dag,\epsilon^\dag)$ form a commutative
monoid':
\begin{gather}
  \delta^\dag \circ (\delta^\dag \otimes \id{Q})
  =    \delta^\dag   \circ ( \id{Q} \otimes \delta^\dag) 
  \tag{A}\label{monoid-assoc}
  \\
  \delta^\dag  \circ (\epsilon^\dag \otimes \id{Q}) 
  = \delta^\dag \circ (\id{Q} \otimes \epsilon^\dag)
  = \id{Q} \tag{U} \label{monoid-unit}
  \\
  \delta^\dag = \delta^\dag \circ
  \sigma. \tag{C}\label{monoid-comm}.
\end{gather}
Futher, these jointly satisfy the \emph{isometry} and \emph{frobenius} laws:
\begin{gather}
  \delta^\dag \circ \delta = \id{Q} \tag{I}\label{isometry}
  \\
  (\delta^\dag \otimes \id{Q}) \circ (\id{Q}\otimes \delta)
  = \delta \circ \delta^\dag. \tag{F}\label{frobenius}
\end{gather}

its denotation is the
  2-dimensional Hilbert space $\mathbb{C}^2$

>>>>

%%%%  use this to include graphics in the right place.
\newcommand{\inlinegraphic}[2]{
  %% todo -- make this thing calculate the height 
  %% itself based on a global scaling factor
  \dimendef\grafheight=255\dimendef\grafvshift=254
  \grafheight=#1
  \grafvshift=-0.5\grafheight
  \advance\grafvshift by 0.5ex
  \raisebox{\grafvshift}{\includegraphics[height=\grafheight]{images/#2}\xspace}
}

@@

Few examples of interesting compact closed categories are freely
generated.  In practice we work with categories presented as
generators and \emph{equations}.  For 
representing quantum computations, Coecke and Duncan
\cite{Coecke:2008jo} propose the following collection of generators:
\begin{description}
\item[Objects] the self dual object $Q$;
\item[Arrows] Two families of arrows:
  \begin{align}
  &\epsilon_Z : Q \to I, &\qquad&  \epsilon_X : Q \to I,\\
  &\delta_Z : Q \to Q \otimes Q, && \delta_X : Q \to Q \otimes Q, \\
  &\alpha_Z : Q \to Q &&  \alpha_X : Q \to Q   
  \end{align}
  where $\alpha \in [0,2\pi)$, and in addition $H:Q\to Q$.
\end{description}
We represent these two families, and their adjoints, by colours in our
graphical notation:
\begin{gather*}
  \epsilon_Z = \inlinegraphic{1.5em}{epsilon} \qquad
  \delta_Z = \inlinegraphic{1.5em}{delta} \qquad
  \epsilon_Z^\dag = \inlinegraphic{1.5em}{epsilondag} \qquad
  \delta_Z^\dag = \inlinegraphic{1.5em}{deltadag} \qquad
  \alpha_Z = \inlinegraphic{1.5em}{greenalpha} \qquad
\\  
  \epsilon_X = \inlinegraphic{1.5em}{redepsilon} \qquad
  \delta_X = \inlinegraphic{1.5em}{reddelta} \qquad
  \epsilon_X^\dag = \inlinegraphic{1.5em}{redepsilondag} \qquad
  \delta_X^\dag = \inlinegraphic{1.5em}{reddeltadag} \qquad
  \alpha_X = \inlinegraphic{1.5em}{redalpha} \qquad
\end{gather*}
The $H$ arrow is denoted by \inlinegraphic{1.5em}{H}.  The free
compact closed category  is then given by all graphs formed by
composing and tensoring these basic graphs.

To model the behaviour of quantum systems these generators must
satisfy certain equations.

Each family $(\delta, \epsilon, \alpha)$ is governed by a set of
equations defining a forms a \emph{classical structure}
\cite{Coecke2006Quantum-Measure,Coecke2006POVMs-and-Naima} with a
\emph{phase group}.  The requisite equations state that $(\delta,
\epsilon )$ form a cocommutative comonoid:
\begin{gather}
  (\delta \otimes \id{Q}) \circ \delta 
   =   ( \id{Q} \otimes \delta) \circ \delta  
  \tag{A}\label{comonoid-assoc}
  \\
  (\epsilon \otimes \id{Q}) \circ \delta 
  = (\id{Q} \otimes \epsilon)\circ \delta 
  = \id{Q} \tag{U} \label{comonoid-unit}
  \\
  \delta = \sigma \circ \delta. \tag{C}\label{comonoid-comm}
\end{gather}
Their adjoints $(\delta^\dag,\epsilon^\dag)$ form a commutative
monoid':
\begin{gather}
  \delta^\dag \circ (\delta^\dag \otimes \id{Q})
  =    \delta^\dag   \circ ( \id{Q} \otimes \delta^\dag) 
  \tag{A}\label{monoid-assoc}
  \\
  \delta^\dag  \circ (\epsilon^\dag \otimes \id{Q}) 
  = \delta^\dag \circ (\id{Q} \otimes \epsilon^\dag)
  = \id{Q} \tag{U} \label{monoid-unit}
  \\
  \delta^\dag = \delta^\dag \circ
  \sigma. \tag{C}\label{monoid-comm}.
\end{gather}
Futher, these jointly satisfy the \emph{isometry} and \emph{frobenius} laws:
\begin{gather}
  \delta^\dag \circ \delta = \id{Q} \tag{I}\label{isometry}
  \\
  (\delta^\dag \otimes \id{Q}) \circ (\id{Q}\otimes \delta)
  = \delta \circ \delta^\dag. \tag{F}\label{frobenius}
\end{gather}

The need to formalise these equations leads naturally to a rewriting
theory
